---
title: "HotHandAnalysisUpdated"
format: html
editor: visual
name: Cameron An
---

## Reading in Player Totals by Season

```{r}
# Loading packages
library(broom)
library(dplyr)
library(forcats)
library(ggplot2)
library(here)
library(hoopR)
library(knitr)
library(tidyverse)
library(stringr)
library(tidyr)
library(purrr)
library(stringr)
library(stringi)
library(kableExtra)
library(scales)
library(gitcreds)
library(ggridges)
library(glmmTMB)
```

```{r}
# Function to read in all player total data from all seasons

read_and_assign_player_totals <- function(start_year = 2015, end_year = 2024) {
  for (year in start_year:end_year) {
    next_year <- year + 1
    season <- sprintf("%02d-%02d", year %% 100, next_year %% 100)
    file_name <- paste0(season, "PlayerData.csv")
    file_path <- here::here("Data", file_name)

    if (file.exists(file_path)) {
      # For 2000–01 through 2002–03 → normal header
      # For 2003–04 and later → skip first line, use second as header
      if (year < 2003) {
        data <- read.csv(file_path, header = TRUE)
      } else {
        # Read first two lines separately to get proper column names
        header_lines <- readLines(file_path, n = 2)
        header <- strsplit(header_lines[2], ",")[[1]]
        data <- read.csv(file_path, skip = 2, header = FALSE)
        names(data) <- header
      }

      # Assign dynamically to global environment
      var_name <- paste0("playerTotals", season)
      assign(var_name, data, envir = .GlobalEnv)
      message("Loaded: ", var_name)

    } else {
      warning("File not found: ", file_name)
    }
  }
}

read_and_assign_player_totals()
```

## Cleaning Data for Player Selection

Data Cleaning:

```{r}
# removing data sets with "League Average" row, converting stats to numeric variables
clean_player_totals_datasets <- function(start_year = 2015, end_year = 2024) {
  for (year in start_year:end_year) {
    next_year <- year + 1
    season <- sprintf("%02d-%02d", year %% 100, next_year %% 100)
    var_name <- paste0("playerTotals", season)

    if (exists(var_name, envir = .GlobalEnv)) {
      data <- get(var_name, envir = .GlobalEnv)

      # Remove row with "League Average"
      data <- dplyr::filter(data, Player != "League Average")
      data <- dplyr::filter(data, Player != "Player")

      # Convert columns 6–31 to numeric (suppress warnings for non-numeric strings)
      data[, 6:31] <- lapply(data[, 6:31], function(x) as.numeric(as.character(x)))

      # Assign cleaned data back to global environment
      assign(var_name, data, envir = .GlobalEnv)
      message("Cleaned: ", var_name)
    } else {
      warning("Dataset not found: ", var_name)
    }
  }
}

clean_player_totals_datasets()
```

Making sure that only one of each player appears in the totals datasets

```{r}
# function that only keeps combined player totals for players that played for multiple teams
keep_2TM_rows <- function(df) {
  df %>%
    dplyr::group_by(Player) %>%
    dplyr::filter(
      if (any(Team == "2TM")) {
        Team == "2TM"
      } else {
        TRUE
      }
    ) %>%
    dplyr::ungroup()
}

playerTotals_names <- ls(pattern = "^playerTotals")

updated_list <- lapply(playerTotals_names, function(name) {
  df <- get(name)
  keep_2TM_rows(df)
})

names(updated_list) <- playerTotals_names

# write back to global environment
list2env(updated_list, envir = .GlobalEnv)

```

Filtering Data by top 50 shooters in FGA per season

```{r}
create_top50_datasets <- function(start_year = 2015, end_year = 2024) {
  for (year in start_year:end_year) {
    next_year <- year + 1
    season <- sprintf("%02d-%02d", year %% 100, next_year %% 100)

    # Dataset names
    totals_name <- paste0("playerTotals", season)
    new_name <- paste0("players", season, "_top50")

    # Check if the dataset exists
    if (exists(totals_name, envir = .GlobalEnv)) {
      data <- get(totals_name, envir = .GlobalEnv)

      # Ensure required columns exist
      required_cols <- c("Player", "Age", "Team", "Pos", "FG%", "FG", "FGA")
      missing_cols <- setdiff(required_cols, names(data))
      if (length(missing_cols) > 0) {
        warning("Skipping ", totals_name, ": missing columns ", paste(missing_cols, collapse = ", "))
        next
      }

      # Select relevant columns and get top 50 shooters
      top50 <- data |>
        dplyr::select(Player, Age, Team, Pos, `FG%`, FG, FGA) |>
        dplyr::arrange(desc(`FGA`)) |>
        dplyr::slice_head(n = 50)

      # Assign result to global environment
      assign(new_name, top50, envir = .GlobalEnv)
      message("Created: ", new_name)
    } else {
      warning("Dataset not found: ", totals_name)
    }
  }
}

create_top50_datasets()
```

```{r}
# All seasons from 01-02 to 24-25
seasons <- c(
  "15-16", "16-17", "17-18", "18-19", 
  "19-20", "20-21", "21-22", "22-23", "23-24", "24-25"
)

assign_ind_top_players <- function(seasons) {
  for (season in seasons) {
    source_name <- paste0("players", season, "_top50")
    target_name <- paste0("players", gsub("-", "_", season), "Ind")
    
    assign(
      target_name,
      get(source_name)$Player,
      envir = .GlobalEnv
    )
  }
}

assign_ind_top_players(seasons)
```

Cleaning Player Names:

```{r}
players15_16Ind[players15_16Ind == "Nikola Vu?evi?"] <- "Nikola Vucevic"
players16_17Ind[players16_17Ind == "Goran Dragi?"] <- "Goran Dragic"
players16_17Ind[players16_17Ind == "Nikola Vu?evi?"] <- "Nikola Vucevic"
players16_17Ind[players16_17Ind == "Dennis Schr\xf6der"] <- "Dennis Schroder"
players16_17Ind[players16_17Ind == "Kristaps Porzi??is"] <- "Kristaps Porzingis"
players17_18Ind[players17_18Ind == "Nikola Joki?"] <- "Nikola Jokic"
players17_18Ind[players17_18Ind == "Goran Dragi?"] <- "Goran Dragic"
players17_18Ind[players17_18Ind == "Jusuf Nurki?"] <- "Jusuf Nurkic"
players17_18Ind[players17_18Ind == "Dennis Schr\xf6der"] <- "Dennis Schroder"
players18_19Ind[players18_19Ind == "Nikola Vu?evi?"] <- "Nikola Vucevic"
players18_19Ind[players18_19Ind == "Dennis Schr\xf6der"] <- "Dennis Schroder"
players18_19Ind[players18_19Ind == "Luka Don?i?"] <- "Luka Doncic"
players18_19Ind[players18_19Ind == "Bojan Bogdanovi?"] <- "Bojan Bogdanovic"
players18_19Ind[players18_19Ind == "Nikola Joki?"] <- "Nikola Jokic"
players19_20Ind[players19_20Ind == "Nikola Joki?"] <- "Nikola Jokic"
players19_20Ind[players19_20Ind == "Luka Don?i?"] <- "Luka Doncic"
players19_20Ind[players19_20Ind == "Bojan Bogdanovi?"] <- "Bojan Bogdanovic"
players19_20Ind[players19_20Ind == "Kristaps Porzi??is"] <- "Kristaps Porzingis"
players19_20Ind[players19_20Ind == "Dennis Schr\xf6der"] <- "Dennis Schroder"
players19_20Ind[players19_20Ind == "Nikola Vu?evi?"] <- "Nikola Vucevic"
players19_20Ind[players19_20Ind == "Nikola Vu?evi?"] <- "Nikola Vucevic"
players20_21Ind[players20_21Ind == "Nikola Vu?evi?"] <- "Nikola Vucevic"
players20_21Ind[players20_21Ind == "Nikola Joki?"] <- "Nikola Jokic"
players20_21Ind[players20_21Ind == "Luka Don?i?"] <- "Luka Doncic"
players20_21Ind[players20_21Ind == "Bojan Bogdanovi?"] <- "Bojan Bogdanovic"
players21_22Ind[players21_22Ind == "Luka Don?i?"] <- "Luka Doncic"
players21_22Ind[players21_22Ind == "Nikola Joki?"] <- "Nikola Jokic"
players21_22Ind[players21_22Ind == "Nikola Vu?evi?"] <- "Nikola Vucevic"
players22_23Ind[players22_23Ind == "Luka Don?i?"] <- "Luka Doncic"
players22_23Ind[players22_23Ind == "Nikola Joki?"] <- "Nikola Jokic"
players22_23Ind[players22_23Ind == "Nikola Vu?evi?"] <- "Nikola Vucevic"
players22_23Ind[players22_23Ind == "Kristaps Porzi??is"] <- "Kristaps Porzingis"
players23_24Ind[players23_24Ind == "Luka Don?i?"] <- "Luka Doncic"
players23_24Ind[players23_24Ind == "Nikola Vu?evi?"] <- "Nikola Vucevic"
players23_24Ind[players23_24Ind == "Nikola Joki?"] <- "Nikola Jokic"
players23_24Ind[players23_24Ind == "Bogdan Bogdanovi?"] <- "Bogdan Bogdanovic"
players24_25Ind[players24_25Ind == "Nikola Joki?"] <- "Nikola Jokic"
players24_25Ind[players24_25Ind == "Alperen ?eng\xfcn"] <- "Alperen Sengun"
players24_25Ind[players24_25Ind == "Luka Don?i?"] <- "Luka Doncic"
players24_25Ind[players24_25Ind == "Nikola Vu?evi?"] <- "Nikola Vucevic"

players15_16Ind[players15_16Ind == "CJ McCollum"] <- "C.J. McCollum"
players17_18Ind[players17_18Ind == "T.J. Warren"] <- "TJ Warren"

```

## Reading in Event Data

```{r}
load_all_nba_pbp <- function(start_season = 2016, end_season = 2025) {

  for (yr in start_season:end_season) {
    message(paste("Loading season", yr, "..."))
    pbp <- hoopR::load_nba_pbp(season = yr)
    obj_name <- paste0("nba_pbp", yr)
    assign(obj_name, pbp, envir = .GlobalEnv)
    message(paste("Saved as", obj_name))
  }
}
```

```{r}
load_all_nba_pbp(2016, 2025)
```

Filtering based on regular season and only shooting plays

```{r}
for (yr in 2016:2025) {

  obj_name <- paste0("nba_pbp", yr)

  if (exists(obj_name)) {

    df <- get(obj_name)

    df <- df %>%
      filter(season_type == 2, shooting_play == TRUE)

    assign(obj_name, df, envir = .GlobalEnv)

    message(paste("Filtered:", obj_name))
  }
}
```

Getting player IDs

```{r}
# Creating dataset to help with gathering player IDs
id_lookup_24_25 <- nba_pbp2025 %>%
  mutate(
    text = str_remove(text, "\\(.*")
  ) %>%
  filter(
    shooting_play == TRUE,
    str_detect(text, paste(players24_25Ind, collapse = "|"))
  ) %>%
  mutate(
    player = str_extract(text, paste(players24_25Ind, collapse = "|"))
  ) %>%
  select(player, athlete_id_1) %>%
  #distinct() %>%
  arrange(player)

```

```{r}
player_ids_25 <- id_lookup_24_25 %>%
  filter(!is.na(athlete_id_1)) %>%
  count(player, athlete_id_1, name = "n") %>%
  group_by(player) %>%
  slice_max(n, with_ties = FALSE) %>%
  ungroup() %>%
  select(player, athlete_id_1)
```

```{r}
make_player_ids <- function(season_year, player_vector) {

  pbp_name <- paste0("nba_pbp", season_year)
  stopifnot(exists(pbp_name))

  pbp <- get(pbp_name)

  id_lookup <- pbp %>%
    mutate(text = str_remove(text, "\\(.*")) %>%
    filter(
      shooting_play == TRUE,
      str_detect(text, paste(player_vector, collapse = "|"))
    ) %>%
    mutate(
      player = str_extract(text, paste(player_vector, collapse = "|"))
    ) %>%
    select(player, athlete_id_1)

  player_ids <- id_lookup %>%
    filter(!is.na(athlete_id_1)) %>%
    count(player, athlete_id_1, name = "n") %>%
    group_by(player) %>%
    slice_max(n, with_ties = FALSE) %>%
    ungroup() %>%
    select(player, athlete_id_1)

  return(player_ids)
}


```

```{r}
season_map <- list(
  "15_16" = 2016,
  "16_17" = 2017,
  "17_18" = 2018,
  "18_19" = 2019,
  "19_20" = 2020,
  "20_21" = 2021,
  "21_22" = 2022,
  "22_23" = 2023,
  "23_24" = 2024,
  "24_25" = 2025
)

for (s in names(season_map)) {

  players_vec <- get(paste0("players", s, "Ind"))

  ids <- make_player_ids(season_map[[s]], players_vec)

  assign(paste0("player_ids_", s), ids, envir = .GlobalEnv)

  message(paste("Created player_ids_", s))
}

```

## Filtering PBP data so it only contains shot attempts from my sample of players

```{r}
# Map pbp years to player_ids seasons
pbp_years <- 2016:2025
player_seasons <- c("15_16","16_17","17_18","18_19","19_20",
                     "20_21","21_22","22_23","23_24","24_25")

for(i in seq_along(pbp_years)) {

  pbp_name <- paste0("nba_pbp", pbp_years[i])
  ids_name <- paste0("player_ids_", player_seasons[i])

  pbp <- get(pbp_name)
  ids <- get(ids_name)

  pbp_filtered <- pbp %>%
    filter(athlete_id_1 %in% ids$athlete_id_1)

  assign(pbp_name, pbp_filtered, envir = .GlobalEnv)
}

```

Deleting free throw observations

```{r}
pbp_years <- 2016:2025

for (yr in pbp_years) {

  pbp_name <- paste0("nba_pbp", yr)

  if (exists(pbp_name)) {

    pbp <- get(pbp_name)

    pbp_filtered <- pbp %>%
      filter(!str_detect(type_text, "Free Throw"))

    assign(pbp_name, pbp_filtered, envir = .GlobalEnv)

    message(paste("Removed Free Throws from", pbp_name))
  }
}

```

## Background Plots

Reading in Season Avg. Data

```{r}
SeasonAvgs <- read.csv(here::here("Data",
                                    "OverallAveragesBySeason.csv"
                                    )
                              )

SeasonAvgs <- na.omit(SeasonAvgs)

seasons_full <-c("2015-16", "2016-17", "2017-18", "2018-19", "2019-20", 
                 "2020-21", "2021-22", "2022-23", "2023-24", "2024-25")

SeasonAvgs <- SeasonAvgs |>
  filter(Season %in% seasons_full) |>
  mutate(
    # Extract last two digits after the dash
    end_two = str_sub(Season, 6, 7),
    
    # Convert to numeric year (assume all seasons end after 1999)
    season_end = 2000 + as.numeric(end_two)
  )
```

Reading in Season Totals Data

```{r}
SeasonTotals <- read.csv(here::here("Data",
                                    "TotalsBySeason.csv"
                                    )
                              )

SeasonTotals <- na.omit(SeasonTotals)

SeasonTotals <- SeasonTotals %>%
  filter(Season %in% seasons_full) |>
  mutate(
    # Extract last two digits after the dash
    end_two = str_sub(Season, 6, 7),
    
    # Convert to numeric year (assume all seasons end after 1999)
    season_end =
      2000 + as.numeric(end_two)
  )
```

Season vs. League Average FG%

```{r}
ggplot(SeasonAvgs, aes(x = season_end, y = FG.)) +
  geom_point() +
  geom_line(alpha = 0.4) +
  geom_smooth(
    method = "loess",
    se = F,
    linewidth = 1.2
  ) + 
  labs(
    x = "Season (End Year)",
    y = "FG%",
    title = "League-Average Field Goal Percentage by Season",
    subtitle = "Season is denoted as end year of season (e.g. 2025 = 2024-25)"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = 2016:2025) +
  scale_y_continuous(labels = function(x) sprintf("%.1f", x * 100))

```

Season vs. League Average FGA per Game

```{r}
ggplot(SeasonAvgs, aes(x = season_end, y = FGA)) +
  geom_point() +
  geom_line(alpha = 0.4) +
  geom_smooth(
    method = "loess",
    se = F,
    linewidth = 1.2
  ) + 
  labs(
    x = "Season (End Year)",
    y = "FGA per Game",
    title = "League-Average Field Goal Attempts per Game by Season"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = 2016:2025) +
  scale_y_continuous(labels = function(x) sprintf("%.1f", x))

```

Season vs. Pace per Game

```{r}
ggplot(SeasonAvgs, aes(x = season_end, y = Pace)) +
  geom_point() +
  geom_line(alpha = 0.4) +
  geom_smooth(
    method = "loess",
    se = F,
    linewidth = 1.2
  ) + 
  labs(
    x = "Season (End Year)",
    y = "Pace",
    title = "League-Average Pace by Season",
    subtitle = "Pace: average number of possessions a team uses in a 48-minute game"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = 2016:2025) +
  scale_y_continuous(labels = function(x) sprintf("%.1f", x))

```

## Opportunity Analysis

Obtaining shot sequences by game for 2024-25 season (testing)

```{r}
shot_sequences <- nba_pbp2025 %>%
  filter(!is.na(athlete_id_1)) %>%
  arrange(game_id, athlete_id_1, desc(start_game_seconds_remaining)) %>%
  left_join(player_ids_24_25, by = "athlete_id_1") %>%
  group_by(player, athlete_id_1, game_id) %>%
  summarise(
    shots = list(as.integer(scoring_play)),
    .groups = "drop"
  )

```

Function that creates game shot sequences

```{r}
# creating game shot sequences
make_shot_sequences <- function(pbp_data, player_ids) {

  seq_df <- pbp_data %>%
    filter(!is.na(athlete_id_1)) %>%
    arrange(game_id, athlete_id_1, desc(start_game_seconds_remaining)) %>%
    left_join(player_ids, by = "athlete_id_1") %>%
    group_by(player, athlete_id_1, game_id) %>%
    summarise(
      shots = list(as.integer(scoring_play)),
      .groups = "drop"
    )

  return(seq_df)
}

```

```{r}
# running make_shot_sequences function for all seasons
pbp_years <- 2016:2025
player_ids_list <- list(
  "2016" = player_ids_15_16,
  "2017" = player_ids_16_17,
  "2018" = player_ids_17_18,
  "2019" = player_ids_18_19,
  "2020" = player_ids_19_20,
  "2021" = player_ids_20_21,
  "2022" = player_ids_21_22,
  "2023" = player_ids_22_23,
  "2024" = player_ids_23_24,
  "2025" = player_ids_24_25
)

shot_sequences_all <- list()

for (yr in pbp_years) {

  pbp <- get(paste0("nba_pbp", yr))
  player_ids <- player_ids_list[[as.character(yr)]]

  seq_df <- make_shot_sequences(pbp, player_ids)

  shot_sequences_all[[as.character(yr)]] <- seq_df

  message(paste("Created shot sequences for season", yr))
}
```

Adding Number of Opportunities and Number of Opportunity Makes

```{r}
get_opportunity_counts <- function(shots) {
  # shots: numeric vector of 1s (make) and 0s (miss)
  
  if(length(shots) < 4) return(c(n_opportunities = 0, n_made = 0))
  
  # Find indices of shots after 3 consecutive makes
  is_opportunity <- sapply(4:length(shots), function(i) {
    all(shots[(i-3):(i-1)] == 1)
  })
  
  opportunities <- shots[4:length(shots)][is_opportunity]
  
  n_opportunities <- length(opportunities)
  n_opportunity_makes <- sum(opportunities)
  
  return(c(n_opportunities = n_opportunities, n_opportunity_makes = n_opportunity_makes))
}

```

Testing for 2024-25

```{r}
game_summary_2025 <- shot_sequences_all[["2025"]] %>%
  rowwise() %>%
  mutate(
    counts = list(get_opportunity_counts(shots)),
    n_opportunities = counts["n_opportunities"],
    n_made = counts["n_made"]
  ) %>%
  ungroup() %>%
  select(player, athlete_id_1, game_id, n_opportunities, n_made)
```

Creating function that loops through all seasons to create Opportunities

```{r}
# Loop over all seasons
for (yr in names(shot_sequences_all)) {

  seq_df <- shot_sequences_all[[yr]]

  summary_df <- seq_df %>%
    rowwise() %>%
    mutate(
      counts = list(get_opportunity_counts(shots)),
      n_opportunities = counts["n_opportunities"],
      n_opportunity_makes = counts["n_opportunity_makes"]
    ) %>%
    ungroup() %>%
    select(player, athlete_id_1, game_id, n_opportunities, n_opportunity_makes)

  # Save as separate dataset for each season
  assign(paste0("game_opportunities_", yr), summary_df, envir = .GlobalEnv)

  message(paste("Created hot-hand game summary for season", yr))
}


```

Creating large dataset with all game opportunities for each season

```{r}
obj_names <- ls(pattern = "^game_opportunities_")

obj_names <- obj_names[sapply(obj_names, exists)]

game_opportunities_all <- mget(obj_names) %>%
  bind_rows(.id = "source") %>%
  mutate(
    season_end = as.integer(str_extract(source, "\\d{4}"))
  ) %>%
  select(-source)
```


Distribution of Hot Hand Opportunities by Season:

```{r}
season_summary_opportunities <- game_opportunities_all %>%
  group_by(season_end) %>%
  summarise(
    mean_opps = mean(n_opportunities, na.rm = TRUE),
    mean_opps_makes = mean(n_opportunity_makes, na.rm = TRUE),
    mean_opps_percentage = sum(n_opportunity_makes, na.rm = TRUE) /
                            sum(n_opportunities, na.rm = TRUE),
    .groups = "drop"
  )
```

League-Average Hot-Hand Opportunities per Game by Season

```{r}
ggplot(season_summary_opportunities, aes(x = season_end, y = mean_opps)) +
  geom_point() +
  geom_line(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_x_continuous(breaks = 2016:2025) +
  labs(
    x = "Season (End Year)",
    y = "Average Hot-Hand Opportunities per Game",
    title = "Average Hot-Hand Opportunities per Game by Season",
    subtitle = "for top 50 shooters per Season (by FGA)"
  ) +
  theme_minimal()

```

League-Average Hot-Hand FG% by Season

```{r}
ggplot(season_summary_opportunities, aes(x = season_end, y = mean_opps_percentage)) +
  geom_point() +
  geom_line(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_x_continuous(breaks = 2016:2025) +
  labs(
    x = "Season (End Year)",
    y = "Hot-Hand FG%",
    title = "Hot-Hand FG% by Season",
    subtitle = "for top 50 shooters per Season (by FGA)"
  ) +
  theme_minimal() + 
  scale_y_continuous(labels = function(x) sprintf("%.1f", x * 100))


```

Creating baseline FG% for top 50 players by season

```{r}
# function to create combined FG% for top 50 players per season
season_fg_percentage <- function(df) {
  total_fg <- sum(df$FG, na.rm = TRUE)
  total_fga <- sum(df$FGA, na.rm = TRUE)
  total_fg / total_fga
}
```

```{r}
# Vector of seasons
seasons <- 2016:2025

# Corresponding dataset names
dataset_names <- c(
  "players15-16_top50", "players16-17_top50", "players17-18_top50",
  "players18-19_top50", "players19-20_top50", "players20-21_top50",
  "players21-22_top50", "players22-23_top50", "players23-24_top50",
  "players24-25_top50"
)

# applying function 
baseline_fg <- sapply(dataset_names, function(x) {
  df <- get(x)
  season_fg_percentage(df)
})
```

```{r}
# adding baseling FG% to season_summary_opportunities
season_summary_opportunities$baseline_fg_percentage <- baseline_fg
```


Difference between Hot-Hand FG% and Baseline FG%

```{r}
# creating difference variable (between hot hand FG% and baseline fg%)
season_summary_opportunities <- season_summary_opportunities %>%
  mutate(diff = mean_opps_percentage - baseline_fg_percentage)
```

```{r}
ggplot(season_summary_opportunities, aes(x = season_end, y = diff)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_point(size = 2) +
  geom_line(alpha = 0.4) +
  geom_smooth(method = "loess", se = T) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title = "Hot-Hand Advantage Over Baseline by Season",
    x = "Season (End Year)",
    y = "Hot-hand 3P% − Baseline 3P%"
  ) +
  theme_minimal() +
  scale_x_continuous(
    breaks = seq(2016, 2025, by = 1),  # tick every year
    labels = seq(2016, 2025, by = 1)   # labels match ticks
  )

```

Creating dataset that analyzes each game shot sequence and reports opp_pct and other_pct

```{r}
# Creating dataset that analyzes each game shot sequence and reports opp_pct and other_pct
shots_all_df <- imap_dfr(shot_sequences_all, function(df, season) {

  df %>%
    mutate(
      season_end = as.integer(season),
      total_shots = map_int(shots, length),
      total_makes = map_int(shots, sum)
    ) %>%
    select(player, athlete_id_1, game_id, season_end,
           total_shots, total_makes)
})

```

```{r}
# combining game data to get all game by game stats ()
combined_games <- game_opportunities_all %>%
  left_join(shots_all_df,
            by = c("player", "athlete_id_1", "game_id", "season_end"))

```

```{r}
# combining game data to get all game by game stats ()
combined_games <- combined_games %>%
  mutate(

    # Opportunity percentage
    opp_pct = ifelse(n_opportunities == 0,
                     NA,
                     n_opportunity_makes / n_opportunities),

    # Other shots
    other_attempts = total_shots - n_opportunities,
    other_makes = total_makes - n_opportunity_makes,

    other_pct = ifelse(other_attempts == 0,
                       NA,
                       other_makes / other_attempts),
    
    # Difference (opp minus other)
    diff = ifelse(is.na(opp_pct) | is.na(other_pct),
                  NA,
                  opp_pct - other_pct)
  )


```

Adding null statistics and p-values for each game

```{r}
set.seed(123)
# Getting null stats using total_shots and total_makes
null_from_counts <- function(total_shots, total_makes,
                             streak_length = 3,
                             n_sims = 1000) {

  if (total_shots <= streak_length) {
    return(tibble(
      null_mean = NA_real_,
      null_median = NA_real_,
      null_sd = NA_real_,
      null_dist = list(NA_real_)
    ))
  }

  null_stats <- replicate(n_sims, {

    sim_vec <- sample(
      c(rep(1, total_makes),
        rep(0, total_shots - total_makes)),
      size = total_shots,
      replace = FALSE
    )

    streak_stats(sim_vec, streak_length)$phat_Sstreak_vs_others
  })

  null_stats <- null_stats[!is.na(null_stats)]

  tibble(
    null_mean   = mean(null_stats),
    null_median = median(null_stats),
    null_sd     = sd(null_stats),
    null_dist   = list(null_stats)
  )
}

```

```{r}
set.seed(123)
# Computes nulls once per unique (total_shots, total_makes) pair
# eliminates many unneccessary iterations
null_lookup <- combined_games %>%
  distinct(total_shots, total_makes) %>%
  mutate(id = row_number()) %>%
  mutate(
    null = map2(total_shots, total_makes,
                ~ null_from_counts(.x, .y,
                                   streak_length = 3,
                                   n_sims = 1000))
  ) %>%
  unnest(null)

```

```{r}
# creating new dataset with null stats and p-values
all_games_analysis <- combined_games %>%
  left_join(null_lookup,
            by = c("total_shots", "total_makes")) %>%
  mutate(

    # p-value using observed diff
    p_value = ifelse(
      is.na(diff),
      NA_real_,
      map2_dbl(null_dist, diff,
               ~ mean(.x >= .y, na.rm = TRUE))
    ),

    z_score = (diff - null_mean) / null_sd
  ) %>%
  select(-null_dist) |>
  mutate(adj_diff = diff - null_mean)

```

Ridge Plot of adj_diff by Season

```{r}
# Compute season means
season_means <- all_games_analysis %>%
  group_by(season_end) %>%
  summarise(mean_adj = mean(adj_diff, na.rm = TRUE))

# Create factor levels for y-axis
season_levels <- factor(sort(unique(all_games_analysis$season_end)))

# Plot
all_games_analysis %>%
  filter(!is.na(adj_diff)) %>%
  ggplot(aes(
    x = adj_diff, 
    y = factor(season_end, levels = rev(levels(season_levels))), # reversed factor
    fill = factor(season_end)
  )) +
  geom_density_ridges(scale = 1.2, alpha = 0.7, color = "black") +
  geom_point(
    data = season_means,
    aes(
      x = mean_adj,
      y = factor(season_end, levels = rev(levels(season_levels)))
    ),
    size = 2
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Difference in FG% on Opportunities vs. Other Shots (Null-Adjusted)", y = "Season",
       title = "Seasonal Differences: FG% on Opportunities vs. Other Shots (Null-Adjusted)") +
  scale_fill_viridis_d() +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) # x-axis as percentages

```

Same as above using z-scores

```{r}
# Compute season means
season_means_zscores <- all_games_analysis %>%
  group_by(season_end) %>%
  summarise(mean_zscore = mean(z_score, na.rm = TRUE))

# Create factor levels for y-axis
season_levels <- factor(sort(unique(all_games_analysis$season_end)))

# Plot
all_games_analysis %>%
  filter(!is.na(z_score)) %>%
  ggplot(aes(
    x = z_score, 
    y = factor(season_end, levels = rev(levels(season_levels))), # reversed factor
    fill = factor(season_end)
  )) +
  geom_density_ridges(scale = 1.2, alpha = 0.7, color = "black") +
  geom_point(
    data = season_means_zscores,
    aes(
      x = mean_zscore,
      y = factor(season_end, levels = rev(levels(season_levels)))
    ),
    size = 2
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Z-Scores", y = "Season",
       title = "Z-Scores of Difference in Proportions (Opp - Other)") +
  scale_fill_viridis_d() 
```

Scatterplot Plot of Adjusted Difference in Proportions by Number of Opportunities by Season

```{r}
all_games_analysis %>%
  filter(!is.na(adj_diff)) %>%
  ggplot(aes(
    x = adj_diff,
    y = n_opportunities,
    color = factor(season_end)
  )) +
  geom_point(alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_viridis_d() +
  labs(
    x = "Adjusted Difference in Proportions (Opportunity − Other)",
    y = "Number of Opportunities",
    color = "Season",
    title = "Game-Level Hot Hand Effect vs Opportunities"
  ) +
  theme_minimal() +
  scale_y_continuous(
    breaks = seq(
      0,
      max(all_games_analysis$n_opportunities, na.rm = TRUE),
      by = 1
    )
  ) 
```

```{r}
# Same as above but flipping axes
all_games_analysis %>%
  filter(!is.na(adj_diff)) %>%
  ggplot(aes(
    x = n_opportunities,
    y = adj_diff,
    color = factor(season_end)
  )) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_viridis_d() +
  labs(
    y = "Adjusted Difference in Proportions (Opportunity − Other)",
    x = "Number of Opportunities",
    color = "Season",
    title = "Game-Level Hot Hand Effect vs Opportunities"
  ) +
  theme_minimal() +
  scale_x_continuous(
    breaks = seq(
      0,
      max(all_games_analysis$n_opportunities, na.rm = TRUE),
      by = 1
    )
  ) 
```

```{r}
# calculating proportion of each 'n_opportunties' value in which adj_diff > 0
opp_positive_prop <- all_games_analysis %>%
  group_by(n_opportunities) %>%
  #filter(
    #n_opportunities > 0,      # exclude 0 opportunities
  #  !is.na(adj_diff)          # ignore NA adj_diff values
  #) %>%
  summarise(
    n_games = n(),                                 
    prop_adj_diff_gt0 = mean(adj_diff > 0, na.rm = T),        # proportion > 0
    .groups = "drop"
  ) %>%
  add_row( # adding row with n_opportunities == 14 since there are no observations in data
    n_opportunities = 14,
    n_games = 0,
    prop_adj_diff_gt0 = NA_real_
  ) %>%
  arrange(n_opportunities) %>%
  mutate(
    prop_adj_diff_gt0 =
      ifelse(is.nan(prop_adj_diff_gt0) | !is.finite(prop_adj_diff_gt0),
             NA_real_,
             prop_adj_diff_gt0)
  )
```

Proportion of Games with Directional Hot Hand by Number of Opportunities

```{r}
opp_positive_prop %>%
  ggplot(aes(
    x = n_opportunities,
    y = prop_adj_diff_gt0
  )) +
  geom_point() +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  labs(
    y = "Proportion",
    x = "Number of Opportunities",
    title = "Proportion of Games with Adjusted Difference Greater than Zero"
  ) +
  theme_minimal() +
  scale_x_continuous(
    breaks = seq(
      0,
      max(all_games_analysis$n_opportunities, na.rm = TRUE),
      by = 1
    )
  ) 
```


#### Shot Distance Analysis

```{r}
add_shot_distance <- function(start_year = 2016, end_year = 2025) {
  
  for (yr in start_year:end_year) {
    
    df_name <- paste0("nba_pbp", yr)
    
    if (exists(df_name, envir = .GlobalEnv)) {
      
      df <- get(df_name, envir = .GlobalEnv)
      
      # Check columns exist
      required_cols <- c("coordinate_x_raw", "coordinate_y_raw")
      
      if (all(required_cols %in% names(df))) {
        
        df <- df %>%
          mutate(
            shot_dist = sqrt(
              (coordinate_x_raw - 25)^2 +
              coordinate_y_raw^2
            )
          )
        
        assign(df_name, df, envir = .GlobalEnv)
        
        message("Updated: ", df_name)
        
      } else {
        message(df_name, " missing coordinate columns — skipped.")
      }
      
    } else {
      message(df_name, " not found — skipped.")
    }
  }
}
```

```{r}
add_shot_distance()
```


```{r}
# New function that stores shots along with shot_dist
make_shot_sequences_withdist <- function(pbp_data, player_ids) {

  seq_df <- pbp_data %>%
    filter(!is.na(athlete_id_1)) %>%
    arrange(game_id,
            athlete_id_1,
            desc(start_game_seconds_remaining)) %>%
    left_join(player_ids, by = "athlete_id_1") %>%
    group_by(player, athlete_id_1, game_id) %>%
    summarise(

      shots = list(as.integer(scoring_play)),

      shot_dist = list(shot_dist),

      .groups = "drop"
    )

  return(seq_df)
}
```

```{r}
# running make_shot_sequences_withdist function for all seasons
pbp_years <- 2016:2025
player_ids_list <- list(
  "2016" = player_ids_15_16,
  "2017" = player_ids_16_17,
  "2018" = player_ids_17_18,
  "2019" = player_ids_18_19,
  "2020" = player_ids_19_20,
  "2021" = player_ids_20_21,
  "2022" = player_ids_21_22,
  "2023" = player_ids_22_23,
  "2024" = player_ids_23_24,
  "2025" = player_ids_24_25
)

shot_sequences_all_withdist <- list()

for (yr in pbp_years) {

  pbp <- get(paste0("nba_pbp", yr))
  player_ids <- player_ids_list[[as.character(yr)]]

  seq_df <- make_shot_sequences_withdist(pbp, player_ids)

  shot_sequences_all_withdist[[as.character(yr)]] <- seq_df

  message(paste("Created shot sequences for season", yr))
}
```



```{r}
# function that computes overall average distance, opportunity average distance, and 
# non-opportunity average distance
get_distance_stats <- function(shots, distances, streak_length = 3){

  n <- length(shots)

  # not enough shots
  if(n <= streak_length){

    return(
      tibble(
        avg_all_dist = NA_real_,
        avg_opp_dist = NA_real_,
        avg_other_dist = NA_real_
      )
    )
  }

  # identify opportunity shots
  opp_indicator <- sapply(
    (streak_length + 1):n,
    function(i){
      all(shots[(i - streak_length):(i - 1)] == 1)
    }
  )

  opp_index <- which(opp_indicator) + streak_length

  # overall average
  avg_all <- mean(distances, na.rm = TRUE)

  # no opportunities case
  if(length(opp_index) == 0){

    return(
      tibble(
        avg_all_dist = avg_all,
        avg_opp_dist = NA_real_,
        avg_other_dist = avg_all
      )
    )
  }

  avg_opp <- mean(distances[opp_index], na.rm = TRUE)

  other_index <- setdiff(seq_len(n), opp_index)

  avg_other <- mean(distances[other_index], na.rm = TRUE)

  tibble(
    avg_all_dist = avg_all,
    avg_opp_dist = avg_opp,
    avg_other_dist = avg_other
  )
}
```

```{r}
# Apply to all games
distance_summary_all <- lapply(

  names(shot_sequences_all_withdist),

  function(yr){

    df <- shot_sequences_all_withdist[[yr]]

    df %>%

      rowwise() %>%

      mutate(

        dist_stats = list(
          get_distance_stats(
            shots,
            shot_dist,
            streak_length = 3
          )
        )

      ) %>%

      tidyr::unnest(dist_stats) %>%

      mutate(season_end = as.integer(yr)) %>%

      ungroup()

  }

)
```

Combine:

```{r}
# Combine
distance_summary_all <- dplyr::bind_rows(distance_summary_all)
```

Overall League Trend:

```{r}
distance_summary_all %>%
  summarise(

    # overall average distance
    league_all = mean(avg_all_dist, na.rm=TRUE),
    
    # opportunity average distance
    league_opp =
    mean(avg_opp_dist, na.rm=TRUE),
    
    # non-opportunity average distance
    league_other =
    mean(avg_other_dist, na.rm=TRUE)
    
  )
```

Player-Level Trend

```{r}
distance_summary_all %>%
  group_by(season_end, player) %>%
  summarise(

    opp_dist =
    mean(avg_opp_dist, na.rm=TRUE),
    
    other_dist =
    mean(avg_other_dist, na.rm=TRUE)
    
  )
```

Save above table to dataset:

```{r}
player_season_distance <- distance_summary_all %>%
  group_by(season_end, player) %>%
  summarise(
    opp_dist = mean(avg_opp_dist, na.rm = TRUE),
    other_dist = mean(avg_other_dist, na.rm = TRUE),
    .groups = "drop"
  )

player_season_distance <- player_season_distance %>%
  mutate(dist_diff = opp_dist - other_dist)
```

```{r}
# Mean of distance difference
mean(player_season_distance$dist_diff, na.rm = TRUE)
# Median of distance difference
median(player_season_distance$dist_diff, na.rm = TRUE)
# One-Sample T-Test of Difference in Differences
t.test(player_season_distance$dist_diff, alternative = "greater")
```

T-Test is statistically significant. Players tend to shoot from farther distances on opportunity shots.


Histogram of Differences in Shot Distance for Opportunity Shots and Non-Opportunity Shots
```{r}
ggplot(player_season_distance,
       aes(x = dist_diff)) +

  geom_histogram(
    bins = 40,
    fill = "#2C7FB8",   # nice blue
    color = "white",
    alpha = 0.9
  ) +

  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 1
  ) +

  labs(
    x = "Opportunity Shot Distance − Other Shot Distance (Ft.)",
    y = "Number of Player-Seasons",
    title = "Average Difference in Shot Distances for Opportunity Shots and Non-Opportunity Shots"
  ) +

  theme_minimal()
```

Boxplots of Differences in Shot Distance for Opportunity Shots and Non-Opportunity Shots by Season

```{r}
ggplot(player_season_distance,
       aes(x = factor(season_end),
           y = dist_diff)) +

  geom_boxplot(fill = "#74A9CF", alpha = .8) +

  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 1
  ) +

  labs(
    x = "Season (End Year)",
    y = "Opportunity Distance − Other Distance (Ft.)",
    title = "Difference in Shot Distances by Season"
  ) +

  theme_minimal()
```

```{r}
player_season_distance <- player_season_distance %>%
  arrange(desc(dist_diff))

head(player_season_distance)
```

```{r}
ggplot(player_season_distance,
       aes(x = other_dist,
           y = opp_dist,
           color = factor(season_end))) +

 geom_point(alpha = .6) +

 geom_abline(
   slope = 1,
   intercept = 0,
   linetype = "dashed"
 ) +

 theme_minimal() +
  scale_color_viridis_d() +
  labs(
    x = "Other Distance (Ft.)",
    y = "Opportunity Distance (Ft.)",
    title = "Average Opportunity Distance vs. Non-Opportunity Distance by Season",
    color = "Season (End Year)"
  )
```


## Hierarchical Model

#### Distance-Adjusted Opportunity Analysis

```{r}
# function to create shot-level data (one row per shot)
make_shot_level <- function(season_df, season){

  season_df %>%
    mutate(

      shot_index =
        map(shots, seq_along),

      is_opportunity =
        map(shots, function(sh){

          n <- length(sh)

          if(n < 4)
            return(rep(FALSE, n))

          opp <- rep(FALSE, n)

          for(i in 4:n){

            if(all(sh[(i-3):(i-1)] == 1))
              opp[i] <- TRUE

          }

          opp
        })

    ) %>%

    mutate(

      shot_data =
        pmap(
          list(
            player,
            athlete_id_1,
            game_id,
            shots,
            shot_dist,
            is_opportunity
          ),

          function(player,
                   athlete_id_1,
                   game_id,
                   shots,
                   shot_dist,
                   is_opportunity){

            tibble(

              player = player,
              athlete_id_1 = athlete_id_1,
              game_id = game_id,
              season_end = season,

              make = shots,

              shot_dist = shot_dist,

              is_opportunity =
                as.integer(is_opportunity)

            )

          }

        )

    ) %>%

    select(shot_data) %>%

    unnest(shot_data)

}
```

```{r}
# creating shot-level-data (one row per shot)
shot_level_data <-
  imap_dfr(
    shot_sequences_all_withdist,
    make_shot_level
  )
```

```{r}
# Adding shot categories to dataset based on shot distance
shot_level_data <- shot_level_data %>%
  mutate(
    dist_category = case_when(
      shot_dist <= 8 ~ "Close",
      shot_dist <= 22 ~ "Midrange",
      TRUE ~ "ThreePoint"
    )
  )

shot_level_data$dist_category <-
  factor(
    shot_level_data$dist_category,
    levels = c("Close","Midrange","ThreePoint")
  )
```

```{r}
# Aggregate the shots
aggregated_shot_data <- shot_level_data %>%
  group_by(
    player, athlete_id_1, season_end, dist_category, is_opportunity
    ) %>%
  summarise(
    attempts = n(),
    makes = sum(make),
    .makegroups = "drop"
    )
```

```{r}
hierarchical_model <- glmmTMB(
  cbind(makes, attempts - makes) ~
    is_opportunity * dist_category +
    (1 + is_opportunity || player) + #+ # || removes correlation estimation
    (1 | player:season_end), # treats 2016 Curry different than 2024 Curry
    #(1 | player) +
    #(1 | season_end),
  family = binomial,
  data = aggregated_shot_data
)
```

```{r}
summary(hierarchical_model)
```

```{r}
# player random effects
player_effects <- ranef(hierarchical_model)$cond$player %>%
  as.data.frame() %>%
  #rownames_to_column(var = "player") %>%
  arrange(desc(is_opportunity))
```


## Season-Long Trends

```{r}
make_season_shot_sequences <- function(pbp_data, player_ids) {
  
  seq_df <- pbp_data %>%
    filter(!is.na(athlete_id_1), athlete_id_1 %in% player_ids$athlete_id_1) %>%
    arrange(athlete_id_1, wallclock, desc(start_game_seconds_remaining)) %>%
    left_join(player_ids, by = "athlete_id_1") %>%
    group_by(player, athlete_id_1) %>%           # only group by player for season
    summarise(
      shots = list(as.integer(scoring_play)),   # TRUE → 1, FALSE → 0
      .groups = "drop"
    )
  
  return(seq_df)
}

```

```{r}
pbp_years <- 2016:2025
player_ids_list <- list(
  "2016" = player_ids_15_16,
  "2017" = player_ids_16_17,
  "2018" = player_ids_17_18,
  "2019" = player_ids_18_19,
  "2020" = player_ids_19_20,
  "2021" = player_ids_20_21,
  "2022" = player_ids_21_22,
  "2023" = player_ids_22_23,
  "2024" = player_ids_23_24,
  "2025" = player_ids_24_25
)

season_shot_sequences <- list()

for (yr in pbp_years) {

  pbp <- get(paste0("nba_pbp", yr))
  player_ids <- player_ids_list[[as.character(yr)]]

  seq_df <- make_season_shot_sequences(pbp, player_ids)

  season_shot_sequences[[as.character(yr)]] <- seq_df

  message(paste("Created season shot sequences for season", yr))
}

```

### Streak Statistics

```{r}
# function that returns basic streak statistics given a sequence
streak_stats = function(results_sequence, streak_length){
  n_trials = length(results_sequence)
  n_success = sum(results_sequence)
  
  run_lengths = rle(results_sequence)$lengths
  if_run_success = rle(results_sequence)$values
  n_runs = length(run_lengths)
  
  # runs of successes
  longest_run_success = max(run_lengths*if_run_success)
  n_runs_success = sum(if_run_success)
  
  # successes after streaks of Successes
  run_lengths_adj = run_lengths - streak_length + 1
  run_values_adj_S =
    if_run_success *
    (!((if_run_success == 1) &
         (run_lengths < streak_length)))
  n_after_Sstreak = sum(run_lengths_adj * run_values_adj_S) -
    run_values_adj_S[length(run_values_adj_S)]
  n_success_after_Sstreak = sum((run_lengths - streak_length) * run_values_adj_S)
  
  # runs of failures
  longest_run_failure = max(run_lengths*(1-if_run_success))
  n_runs_failure = sum(1-if_run_success)
  
  # successes after streaks of Failures
  run_values_adj_F =
    (1-if_run_success) *
    (!((if_run_success == 0) &
         (run_lengths < streak_length)))
  n_after_Fstreak = sum(run_lengths_adj * run_values_adj_F) -
    run_values_adj_F[length(run_values_adj_F)]
  n_success_after_Fstreak = n_after_Fstreak -
    sum((run_lengths - streak_length) * run_values_adj_F)
  
  # Proportion of hits on trials after streaks
  phat_after_Sstreak = n_success_after_Sstreak / n_after_Sstreak
  
  # Difference in proportion of hits (trials after streaks - all other trials)
  phat_after_no_Sstreak = (n_success - n_success_after_Sstreak)/
    (n_trials - n_after_Sstreak)
  phat_Sstreak_vs_others = phat_after_Sstreak - phat_after_no_Sstreak
  
  # Difference in proportion of hits (trials after hit streaks - trials after miss streaks)
  phat_after_Fstreak = n_success_after_Fstreak / n_after_Fstreak
  phat_Sstreak_vs_Fstreak = phat_after_Sstreak - phat_after_Fstreak
  
  # Hit streak frequency - proportion of trials that occur after hit streaks
  Sstreak_frequency  = n_after_Sstreak / (n_trials - streak_length)
  
  # collect the streak stats for the results sequence
  return(data.frame(phat_after_Sstreak,
                    phat_Sstreak_vs_others,
                    phat_Sstreak_vs_Fstreak,
                    Sstreak_frequency,
                    longest_run_success,
                    n_runs
                    #                     n_runs_success,
                    #                     phat_after_Fstreak,
                    #                     n_runs_failure,
                    #                     longest_run_failure,
  ))
}
```

Function that loops through all season-long shot vectors and analyzes them

```{r}
analyze_all_shot_vectors <- function(all_shot_vectors, streak_length = 3) {
  
  results <- list()
  idx <- 1
  
  for (season in names(all_shot_vectors)) {
    for (player in names(all_shot_vectors[[season]])) {
      
      shots <- all_shot_vectors[[season]][[player]]
      
      # Skip sequences too short for streak analysis
      if (length(shots) <= streak_length) next
      
      stats <- streak_stats(shots, streak_length)
      
      results[[idx]] <- cbind(
        season = season,
        player = player,
        n_shots = length(shots),
        stats
      )
      
      idx <- idx + 1
    }
  }
  
  do.call(rbind, results)
}
```

Converting season_shot_sequences object to format that is usable by the function

```{r}
all_season_shot_vectors <- lapply(season_shot_sequences, function(df) {
  
  # df is one season's data frame
  
  shots_list <- df$shots
  names(shots_list) <- df$player
  
  shots_list
})

```

Using above function

```{r}
season_level_results <- analyze_all_shot_vectors(
  all_season_shot_vectors,
  streak_length = 3
)
```

Calculating p-values for all season shot vectors using phat_Streak_vs_others

```{r}
hot_hand_pvalue <- function(
  shot_vec,
  observed_phat,
  streak_length = 3,
  n_sims = 5000
) {
  
  n_trials <- length(shot_vec)
  n_success <- sum(shot_vec)
  
  # Edge cases: no possible streaks
  if (n_trials <= streak_length || is.na(observed_phat)) {
    return(NA_real_)
  }
  
  # Simulate null sequences with same shots & makes
  sim_stats <- replicate(n_sims, {
    
    sim_vec <- sample(
      c(rep(1, n_success), rep(0, n_trials - n_success)),
      size = n_trials,
      replace = FALSE
    )
    
    streak_stats(sim_vec, streak_length)$phat_Sstreak_vs_others
  })
  
  sim_stats <- sim_stats[!is.na(sim_stats)]
  
  # One-sided p-value
  mean(sim_stats >= observed_phat)
}

```

Applying to all player-seasons:

```{r}
library(dplyr)
library(purrr)

season_level_results <- season_level_results %>%
  rowwise() %>%
  mutate(
    pvalue = {
      
      season_chr <- as.character(season)   # e.g. "2002"
      player_chr <- player
      
      shot_vec <- all_season_shot_vectors[[season_chr]][[player_chr]]
      
      if (is.null(shot_vec)) {
        NA_real_
      } else {
        hot_hand_pvalue(
          shot_vec = shot_vec,
          observed_phat = phat_Sstreak_vs_others,
          streak_length = 3,
          n_sims = 5000
        )
      }
    }
  ) %>%
  ungroup()

```

```{r}
# Expect 5% -> 0.05(500) = 25 to have significant p-values
sum(season_level_results$pvalue < 0.05, na.rm = TRUE)
```

Distribution of Hot Hand p-values

```{r}
ggplot(season_level_results, aes(x = pvalue)) +
  geom_density(fill = "steelblue", alpha = 0.4) +
  geom_vline(xintercept = 0.05, linetype = "dashed") +
  labs(
    title = "Density of p-values",
    x = "p-value",
    y = "Density"
  ) +
  theme_minimal()
```

ECDF plot

```{r}
ggplot(season_level_results, aes(pvalue)) +
  stat_ecdf(size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Empirical CDF of Hot Hand P-values",
    x = "P-value",
    y = "Cumulative proportion"
  ) +
  theme_minimal()

```

Interpretation

-   Dashed line = perfect uniform null

-   Curve above line near 0 = excess small p-values

QQ-Plot

```{r}
observed <- sort(season_level_results$pvalue)
expected <- ppoints(length(observed))

qq_df <- data.frame(expected, observed)

ggplot(qq_df, aes(expected, observed)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "QQ Plot of Hot Hand P-values",
    x = "Expected under Uniform(0,1)",
    y = "Observed p-values"
  ) +
  theme_minimal()
```

Season-by-season p value distributions

```{r}
ggplot(season_level_results,
       aes(x = factor(season), y = pvalue)) +
  geom_boxplot(outlier.alpha = 0.4) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  labs(
    title = "Distribution of Hot Hand P-values by Season",
    x = "Season",
    y = "P-value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45))

```

Benjamini-Hochberg Adjusted P-Values

```{r}
season_level_results <- season_level_results %>%
  mutate(
    pvalue_adj = p.adjust(pvalue, method = "BH")
  )
```

#### Null Distribution Statistics

```{r}
hot_hand_null_stats <- function(
  shot_vec,
  observed_phat,
  streak_length = 3,
  n_sims = 5000
) {

  n_trials  <- length(shot_vec)
  n_success <- sum(shot_vec)

  # Edge cases
  if (n_trials <= streak_length || is.na(observed_phat)) {
    return(
      tibble(
        null_median = NA_real_,
        null_mean   = NA_real_,
        null_sd     = NA_real_,
        p_value     = NA_real_,
        z_score     = NA_real_
      )
    )
  }

  # Simulate null distribution
  null_stats <- replicate(n_sims, {
    sim_vec <- sample(
      c(rep(1, n_success), rep(0, n_trials - n_success)),
      size = n_trials,
      replace = FALSE
    )
    streak_stats(sim_vec, streak_length)$phat_Sstreak_vs_others # $phat_after_Sstreak
  })

  null_stats <- null_stats[!is.na(null_stats)]

  null_mean   <- mean(null_stats)
  null_median <- median(null_stats)
  null_sd     <- sd(null_stats)

  tibble(
    null_median = null_median,
    null_mean   = null_mean,
    null_sd     = null_sd,
    p_value     = mean(null_stats >= observed_phat),
    z_score     = (observed_phat - null_mean) / null_sd
  )
}

```

```{r}
season_level_results <- season_level_results %>%
  mutate(
    diff_in_proportions = phat_Sstreak_vs_others
  )
```

```{r}
library(dplyr)
library(purrr)

season_level_results <- season_level_results %>%
  rowwise() %>%
  mutate(
    null_stats = list({

      season_chr <- as.character(season)
      player_chr <- player
      shot_vec   <- all_season_shot_vectors[[season_chr]][[player_chr]]

      if (is.null(shot_vec)) {
        tibble(
          null_median = NA_real_,
          null_mean   = NA_real_,
          null_sd     = NA_real_,
          p_value     = NA_real_,
          z_score     = NA_real_
        )
      } else {
        hot_hand_null_stats(
          shot_vec       = shot_vec,
          observed_phat  = phat_Sstreak_vs_others,
          streak_length  = 3,
          n_sims         = 5000
        )
      }
    })
  ) %>%
  unnest(null_stats) %>%
  ungroup()

```

```{r}
appendix_table <- season_level_results %>%
  rowwise() %>%
  mutate(
    n_successes = sum(all_season_shot_vectors[[as.character(season)]][[player]]),
    phat_other_trials = phat_after_Sstreak - phat_Sstreak_vs_others
  ) %>%
  ungroup()


appendix_table <- appendix_table %>%
  select(
    player,
    season,
    n_shots,
    n_successes,
    n_runs,
    phat_after_Sstreak,
    phat_other_trials,
    diff_in_proportions,
    null_median,
    null_mean,
    null_sd,
    p_value,
    z_score
  ) %>%
  arrange(p_value)


```

```{r}
appendix_table_top20 <- appendix_table %>%
  slice_head(n = 20)


appendix_table_top20 %>%
  kable(
    format = "html",
    caption = "Top 20 Player–Seasons with the Lowest Permutation-Test p-values (Streak Length = 3)",
    digits = 3
  ) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "hover", "condensed")
  )
```

### k = 1

Using above function

```{r}
season_level_results_k1 <- analyze_all_shot_vectors(
  all_season_shot_vectors,
  streak_length = 1
)
```

Calculating p-values for all season shot vectors using phat_Streak_vs_others

```{r}
library(dplyr)
library(purrr)

season_level_results_k1 <- season_level_results_k1 %>%
  rowwise() %>%
  mutate(
    pvalue = {
      
      season_chr <- as.character(season)   # e.g. "2002"
      player_chr <- player
      
      shot_vec <- all_season_shot_vectors[[season_chr]][[player_chr]]
      
      if (is.null(shot_vec)) {
        NA_real_
      } else {
        hot_hand_pvalue(
          shot_vec = shot_vec,
          observed_phat = phat_Sstreak_vs_others,
          streak_length = 1,
          n_sims = 2000
        )
      }
    }
  ) %>%
  ungroup()

```

```{r}
# Expect 5% -> 0.05(500) = 25 to have significant p-values
sum(season_level_results_k1$pvalue < 0.05, na.rm = TRUE)
```

#### Null Distribution Statistics

```{r}
season_level_results_k1 <- season_level_results_k1 %>%
  mutate(
    diff_in_proportions = phat_Sstreak_vs_others
  )
```

```{r}
library(dplyr)
library(purrr)

season_level_results_k1 <- season_level_results_k1 %>%
  rowwise() %>%
  mutate(
    null_stats = list({

      season_chr <- as.character(season)
      player_chr <- player
      shot_vec   <- all_season_shot_vectors[[season_chr]][[player_chr]]

      if (is.null(shot_vec)) {
        tibble(
          null_median = NA_real_,
          null_mean   = NA_real_,
          null_sd     = NA_real_,
          p_value     = NA_real_,
          z_score     = NA_real_
        )
      } else {
        hot_hand_null_stats(
          shot_vec       = shot_vec,
          observed_phat  = phat_Sstreak_vs_others,
          streak_length  = 1,
          n_sims         = 5000
        )
      }
    })
  ) %>%
  unnest(null_stats) %>%
  ungroup()

```

```{r}
appendix_table_k1 <- season_level_results_k1 %>%
  rowwise() %>%
  mutate(
    n_successes = sum(all_season_shot_vectors[[as.character(season)]][[player]]),
    phat_other_trials = phat_after_Sstreak - phat_Sstreak_vs_others
  ) %>%
  ungroup()


appendix_table_k1 <- appendix_table_k1 %>%
  select(
    player,
    season,
    n_shots,
    n_successes,
    n_runs,
    phat_after_Sstreak,
    phat_other_trials,
    diff_in_proportions,
    null_median,
    null_mean,
    null_sd,
    p_value,
    z_score
  ) %>%
  arrange(p_value)


```
